Template for indicator documentation

### 2023-09-20 NDVI_wetland.Rmd
Small changes in the code to make it run in the book

#### NiN data and geometry for Norway and the five regions
We read the NiN data for this document from a cache at the NINA server...
```{r, echo = F}
# Add NiN data from cache
nin <- readRDS(paste0(here::here(),"/data/cache/nin.RDS"))
```

...but it can also be directly downloaded from Miljødirektoratets kartkatalog (https://kartkatalog.miljodirektoratet.no/Dataset)
```{r, echo = T, eval = F}

url <- "https://nedlasting.miljodirektoratet.no/Miljodata//Naturtyper_nin/FILEGDB/4326/Naturtyper_nin_0000_norge_4326_FILEGDB.zip"
download(url, dest="P:/41201785_okologisk_tilstand_2022_2023/data/Naturtyper_nin_0000_norge_4326_FILEGDB.zip", mode="wb") 
unzip ("P:/41201785_okologisk_tilstand_2022_2023/data/Naturtyper_nin_0000_norge_4326_FILEGDB.zip", 
       exdir = "P:/41201785_okologisk_tilstand_2022_2023/data/nin_data")
st_layers("P:/41201785_okologisk_tilstand_2022_2023/data/nin_data/Naturtyper_nin_0000_norge_4326_FILEGDB.gdb")
nin <- st_read("P:/41201785_okologisk_tilstand_2022_2023/data/nin_data/Naturtyper_nin_0000_norge_4326_FILEGDB.gdb",
                 layer="naturtyper_nin_omr")
```
And we upload maps for Norway and the five regions: Southern, Western, Eastern, Central, and Northern Norway
```{r message=FALSE}
# Import region- og Norgeskart, make sure they have the same geometry as the nin data
nor <- st_read("data/outlineOfNorway_EPSG25833.shp")%>%
  st_as_sf() %>%
  st_transform(crs = st_crs(nin))

reg <- st_read("data/regions.shp")%>%
  st_as_sf() %>%
  st_transform(crs = st_crs(nin))

# change region names to something R-friendly
reg$region
reg$region <- c("Northern Norway","Central Norway","Eastern Norway","Western Norway","Southern Norway")

# combine the Norway and the region maps
regnor <- st_intersection(reg,nor)
rm(reg)
rm(nor)
```

#### NDVI data
All NDVI data are calculated mean NDVI values for each respective NiN polygon from each available Sentinel-2, MODIS, or Landsat image in Google Earth Engine (GEE). 
```{r, echo = F}
# downloading the unprocessed NDVI data from a data-cache et the NINA server
df.s <- readRDS("P:/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/ndvi.Sentinel.raw.RDS")
df.m <- readRDS("P:/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/ndvi.Modis.raw.RDS")
df.l <- readRDS("P:/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/ndvi.Landsat.raw.RDS")
```
##### Sentinel-2 NDVI data
The GEE code can be seen here: https://code.earthengine.google.com/2ceb0c3e03adade9e6f6d0903184b8c4
The image collection contains Sentinel imagery from June, July and August 2015-2022
To not exceed GEE memory limits, the exported files had to be iterated over a grid which resulted in 42 separate csv files. This script merges them and then merges the dataframe to the NiN data.
```{r, echo = T, eval = F}
# Import Sentinel NDVI Data
df.s <- list.files("P:/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/NDVI_data_Sentinel/", pattern = "*.csv", full.names=TRUE) %>%
  map_df(~fread(.))
```

@@

```{r, echo = T, eval = F}
# Import MODIS NDVI Data
# MODIS NDVI is scaled by 0.0001. Mean must be divided by 10000.
df.m<- read.csv("P:\\41201785_okologisk_tilstand_2022_2023\\data\\NDVI_åpenlavland\\NDVI_data_MODIS\\modis_ndvi_ts_2000_2022.csv", )
df.m$mean<-df.m$mean/10000
```

##### Landsat NDVI data
The GEE code can be seen here: https://code.earthengine.google.com/da8a9279238ef26d14be08a43788b6b7
The image collection contains Landsat imagery from June, July and August 1984-2022
To not exceed GEE memory limits, the exported files had to be iterated over a grid which resulted in 42 separate csv files. This script merges them and then merges the dataframe to the NiN data.
```{r, echo = T, eval = F}
# Import Landsat NDVI Data
# Set up conditional file paths
dir <- substr(getwd(), 1,2)

path <- ifelse(dir == "C:", 
               "R:/GeoSpatialData/Habitats_biotopes/Norway_Miljodirektoratet_Naturtyper_nin/Original/versjon20221231/Natur_Naturtyper_nin_norge_med_svalbard_25833/Natur_Naturtyper_NiN_norge_med_svalbard_25833.gdb",
               "/data/R/GeoSpatialData/Habitats_biotopes/Norway_Miljodirektoratet_Naturtyper_nin/Original/versjon20221231/Natur_Naturtyper_nin_norge_med_svalbard_25833/Natur_Naturtyper_NiN_norge_med_svalbard_25833.gdb")

pData <- ifelse(dir == "C:", 
                "P:/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/NDVI_data_Landsat",
                "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/NDVI_data_Landsat")

# Fread doesn't like the weird path to the server version of the P drive
# hence this horrendous work around
files=list.files(pData, pattern = "*.csv", full.names = TRUE)

df_list<-list()# initialise a list of dataframes
# read in a dataframe in each slot of the df_list
for (i in files){
  name <- gsub("-",".",i)
  name <- gsub(".csv","",name)  
  i <- paste(i,sep="")
  df_list[[i]]=assign(name,read.csv(i, header=TRUE))
}

df.l<-bind_rows(df_list, .id = "column_label")
```

### Data handling
#### NiN data
```{r, echo = T, eval = F}
# fixing variable- and ecosystem-names with special characters
colnames(nin)
colnames(nin)[c(3,8,17,26,31,33,34)] <- c("hovedoekosystem","kartleggingsaar","noyaktighet",
                                          "omraadenavn","uk_naertruet","uk_sentraloekosystemfunksjon",
                                          "uk_spesieltdaarligkartlagt")
unique(nin$hovedoekosystem)

nin <- nin %>% mutate(hovedoekosystem = recode(hovedoekosystem, 
                                               "Våtmark" = 'Vaatmark',
                                               "Semi-naturlig mark" = 'Semi_naturlig',
                                               "Naturlig åpne områder i lavlandet" = 'Naturlig-aapent',
                                               "Naturlig åpne områder under skoggrensa" = 'Naturlig_aapent')) %>%
  mutate(validGeo = st_is_valid(SHAPE))

@@

## filter out only wetland data
nin.wetland <- nin %>% 
  filter(hovedoekosystem %in% c('Vaatmark')) %>%
  mutate(id = identifikasjon_lokalid) %>%
  filter(validGeo) %>%
  drop_na(tilstand) %>%
  dplyr::select(id, hovedoekosystem, hovedtype, ninkartleggingsenheter, lokalitetskvalitet, tilstand, tilstand_num, ninbeskrivelsesvariabler, kartleggingsaar)

# merge NiN-data with region
nin.wetland = st_join(nin.wetland, regnor, left = TRUE)
nin.wetland

colnames(nin.wetland)[c(1,10)] <- c("id","region_id")

# check that every nin-polygon still occurs only once
summary(as.factor(nin.wetland$id))
nin.wetland[nin.wetland$id=="NINFP2110008060",]
# this one was assigned to both Western and Central Norway, we drop the latter
nin.wetland <- nin.wetland[!row.names(nin.wetland) %in% '1144.1',]

nin.wetland <- nin.wetland %>% 
  mutate(area_meters_nin = st_area(nin.wetland)
  )
# check and edit the order of regions
levels(nin.wetland$region)
nin.wetland$region <- as.factor(nin.wetland$region)
levels(nin.wetland$region)
nin.wetland$region <- factor(nin.wetland$region, levels = c("Northern Norway","Central Norway","Eastern Norway","Western Norway","Southern Norway"))
levels(nin.wetland$region)

# splitting content in the ninbeskrivelsesvariabler-column into separate rows per variable and then separating that column into two for the variable name and value
nin.wetland2 <- nin.wetland %>% separate_rows(ninbeskrivelsesvariabler, sep=",") %>%
  separate(col = ninbeskrivelsesvariabler,
           into = c("NiN_variable_code", "NiN_variable_value"),
           sep = "_",
           remove=F) %>%
  mutate(NiN_variable_value = as.numeric(NiN_variable_value)) %>%
  mutate(NiN_variable_code = as.factor(NiN_variable_code)) %>%
  subset(select = -ninbeskrivelsesvariabler)
nin.wetland2 <- as.data.frame(nin.wetland2)

```

```{r, echo = F}

#load processed NiN-data from cache
nin.wetland <- readRDS(paste0(here::here(),"/data/cache/nin.wetland.RDS"))
nin.wetland2 <- readRDS(paste0(here::here(),"/data/cache/nin.wetland2.RDS"))

```

#### NDVI data
```{r, results='hide'}
## Sentinel-2
# join nin.wetland & Sentinel NDVI data
SentinelNDVI.wetland <- full_join(nin.wetland, df.s, by="id")
#summary(SentinelNDVI.wetland)
SentinelNDVI.wetland <- SentinelNDVI.wetland %>%
  mutate(hovedoekosystem = as.factor(hovedoekosystem),
         hovedtype = as.factor(hovedtype),
         ninkartleggingsenheter = as.factor(ninkartleggingsenheter), 
         lokalitetskvalitet = as.factor(lokalitetskvalitet),
         tilstand = as.factor(tilstand),
         area_meters = st_area(SentinelNDVI.wetland))
#summary(SentinelNDVI.wetland)
# get rid of NAs (i.e. NDVI cells that were not in wetland polygons)
SentinelNDVI.wetland <- SentinelNDVI.wetland %>% filter(!is.na(hovedtype))
SentinelNDVI.wetland <- SentinelNDVI.wetland %>% filter(!is.na(mean))
#summary(SentinelNDVI.wetland)
# get rid of any nin-polygons smaller than the Sentinel grid cell size (100 sqm)
dim(SentinelNDVI.wetland)
SentinelNDVI.wetland <- SentinelNDVI.wetland %>% filter(as.numeric(area_meters) >= 100)
dim(SentinelNDVI.wetland)
# split date into year, month & day
SentinelNDVI.wetland <- SentinelNDVI.wetland %>%
  dplyr::mutate(year = lubridate::year(date), 
                month = lubridate::month(date), 
                day = lubridate::day(date))
# add column for sub-ecosystem types
SentinelNDVI.wetland <- SentinelNDVI.wetland %>% mutate(subtype = substring(ninkartleggingsenheter, 4),
                                                        subtype = str_remove(subtype, '-'),
                                                        subtype = str_remove(subtype, '-'))
# we are using max NDVI per year in every NiN polygon
SentinelNDVI.wetland <- SentinelNDVI.wetland %>%
  group_by(id, year) %>%
  filter(mean == max(mean, na.rm=TRUE))

@@

## MODIS
# join nin.wetland and Modis NDVI data
ModisNDVI.wetland <- full_join(nin.wetland, df.m, by="id")

#summary(ModisNDVI.wetland)
ModisNDVI.wetland <- ModisNDVI.wetland %>%
  mutate(hovedoekosystem = as.factor(hovedoekosystem),
         hovedtype = as.factor(hovedtype),
         ninkartleggingsenheter = as.factor(ninkartleggingsenheter), 
         lokalitetskvalitet = as.factor(lokalitetskvalitet),
         tilstand = as.factor(tilstand),
         area_meters = st_area(ModisNDVI.wetland))
#summary(ModisNDVI.wetland)
# get rid of NAs (i.e. NDVI cells that were not in wetland polygons)
ModisNDVI.wetland <- ModisNDVI.wetland %>% filter(!is.na(hovedtype))
ModisNDVI.wetland <- ModisNDVI.wetland %>% filter(!is.na(mean))
#summary(ModisNDVI.wetland)
# get rid of any nin-polygons smaller than the Modis grid cell size (62500 sqm)
dim(ModisNDVI.wetland)
ModisNDVI.wetland <- ModisNDVI.wetland %>% filter(as.numeric(area_meters) >= 62500)
dim(ModisNDVI.wetland)
# split date into year, month & day
ModisNDVI.wetland <- ModisNDVI.wetland %>%
  dplyr::mutate(year = lubridate::year(date), 
                month = lubridate::month(date), 
                day = lubridate::day(date))
# we are using max NDVI per year in every NiN polygon
ModisNDVI.wetland <- ModisNDVI.wetland %>%
  group_by(id, year) %>%
  filter(mean == max(mean, na.rm=TRUE))
#summary(ModisNDVI.wetland)

## Landsat
# join nin.wetland & Landsat NDVI data
LandsatNDVI.wetland <- full_join(nin.wetland, df.l, by="id")

#summary(LandsatNDVI.wetland)
LandsatNDVI.wetland <- LandsatNDVI.wetland %>%
  mutate(hovedoekosystem = as.factor(hovedoekosystem),
         hovedtype = as.factor(hovedtype),
         ninkartleggingsenheter = as.factor(ninkartleggingsenheter), 
         lokalitetskvalitet = as.factor(lokalitetskvalitet),
         tilstand = as.factor(tilstand),
         area_meters = st_area(LandsatNDVI.wetland))
#summary(LandsatNDVI.wetland)
# get rid of NAs (i.e. NDVI cells that were not in wetland polygons)
LandsatNDVI.wetland <- LandsatNDVI.wetland %>% filter(!is.na(hovedtype))
LandsatNDVI.wetland <- LandsatNDVI.wetland %>% filter(!is.na(mean))
#summary(LandsatNDVI.wetland)
# get rid of any nin-polygons smaller than the Landsat grid cell size (900 sqm)
dim(LandsatNDVI.wetland)
LandsatNDVI.wetland <- LandsatNDVI.wetland %>% filter(as.numeric(area_meters) >= 900)
dim(LandsatNDVI.wetland)
# split date into year, month & day
LandsatNDVI.wetland <- LandsatNDVI.wetland %>%
  dplyr::mutate(year = lubridate::year(date), 
                month = lubridate::month(date), 
                day = lubridate::day(date))
# we are using max NDVI per year in every NiN polygon
LandsatNDVI.wetland <- LandsatNDVI.wetland %>%
  group_by(id, year) %>%
  filter(mean == max(mean, na.rm=TRUE))

#summary(LandsatNDVI.wetland)
```

@@

# Where are the wetland sites?
```{r}
tm_shape(regnor) +
  tm_fill('GID_0', labels="", title="", legend.show = FALSE) + 
  tm_borders() +
  tm_shape(SentinelNDVI.wetland) +
  tm_dots('hovedtype',midpoint=NA, palette=tmaptools::get_brewer_pal("YlOrRd", 6, plot = FALSE), scale=2.5, legend.show = FALSE) + 
  tm_add_legend(type = "symbol", 
                col = tmaptools::get_brewer_pal("YlOrRd", 6, plot = FALSE),
                labels = c("V1 - minerotrophic mires",
                           "V3 - ombrotrophic bogs",
                           "V4 - cold springs",
                           "V6 - wet snowbeds",
                           "V9 - semi-natural mires",
                           "V10 - semi-natural wet meadows"),
                title = "Major ecosystem type",
                size=3) +
  tm_layout(main.title = "Wetland ecosystems (NiN, Miljødirektoratets instruks)",
            main.title.size=1.2, legend.position = c("right", "bottom"), 
            legend.text.size = 1.3, legend.title.size = 1.4)
```

# How are polygon sizes distributed?
```{r}
summary(SentinelNDVI.wetland$area_meters)
hist(SentinelNDVI.wetland$area_meters,xlim=c(0,20000),breaks=50000)
abline(v=100,lty=2)
```
Wetland polygons stretch a large number of sizes from just above 100 sqm to more than 4 million sqm. 75% of all polygons are below 14264 sqm.

# How does NDVI vary over the years
```{r}
SentinelNDVI.wetland %>%
  ggplot( aes(x=year, y=mean )) + 
  geom_point() +
  facet_grid( tilstand~hovedtype)
```
2022 stands out with the highest NDVI values missing.
WE omit 2022 for this analysis
```{r}
SentinelNDVI.wetland <- SentinelNDVI.wetland %>% filter(year != '2022')
```

# How do NDVI values vary between major and basic ecosystem types for sites in good condition? (using only NDVI data matching NiN-mapping years)
```{r}
SentinelNDVI.wetland %>%
  filter(tilstand == 'God') %>%
  filter(year == kartleggingsaar) %>%
  ggplot(aes(x=hovedtype, y=mean )) + 
  geom_violin() +
  geom_point(size=0.7, shape=16, color="grey")

SentinelNDVI.wetland %>%
  filter(tilstand == 'God') %>%
  filter(year == kartleggingsaar) %>%
  ggplot(aes(x=subtype, y=mean )) + 
  geom_violin() +
  geom_point(size=0.7, shape=16, color="grey") +
  facet_wrap( ~hovedtype) +
  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust=0.2))
```
NDVI varies between major ecosystem types with the highest values in semi-natural wet-meadows (V10), and lowest values in ombrotrophic bogs (V3). The overlaps between the ecosystem types are considerable.
With respect to basic ecosystem types within the major types, NDVI appears to increase with increasing limestone level (C1-C4, C5-C8, and E1-E3 in V1, C1-C3 in the other major types), and also here the overlap is large.

>>>>

#### NiN data and geometry for Norway and the five regions

```{r, echo = F}
# Add NiN data from cache
nin <- readRDS(paste0(here::here(), "/data/cache/nin.RDS"))
```

The NiN data can be directly downloaded from Miljødirektoratets kartkatalog (https://kartkatalog.miljodirektoratet.no/Dataset)

```{r, echo = T, eval = F}
url <- "https://nedlasting.miljodirektoratet.no/Miljodata//Naturtyper_nin/FILEGDB/4326/Naturtyper_nin_0000_norge_4326_FILEGDB.zip"
download(url, dest = "P:/41201785_okologisk_tilstand_2022_2023/data/Naturtyper_nin_0000_norge_4326_FILEGDB.zip", mode = "wb")
unzip("P:/41201785_okologisk_tilstand_2022_2023/data/Naturtyper_nin_0000_norge_4326_FILEGDB.zip",
  exdir = "P:/41201785_okologisk_tilstand_2022_2023/data/nin_data"
)
st_layers("P:/41201785_okologisk_tilstand_2022_2023/data/nin_data/Naturtyper_nin_0000_norge_4326_FILEGDB.gdb")
nin <- st_read("P:/41201785_okologisk_tilstand_2022_2023/data/nin_data/Naturtyper_nin_0000_norge_4326_FILEGDB.gdb",
  layer = "naturtyper_nin_omr"
)
```

And we upload maps for Norway and the five regions: Southern, Western, Eastern, Central, and Northern Norway

```{r message=FALSE}
# Import region- og Norgeskart, make sure they have the same geometry as the nin data
nor <- st_read("data/outlineOfNorway_EPSG25833.shp") %>%
  st_as_sf() %>%
  st_transform(crs = st_crs(nin))

reg <- st_read("data/regions.shp") %>%
  st_as_sf() %>%
  st_transform(crs = st_crs(nin))

# change region names to something R-friendly
reg$region
reg$region <- c("Northern Norway", "Central Norway", "Eastern Norway", "Western Norway", "Southern Norway")

# combine the Norway and the region maps
regnor <- st_intersection(reg, nor)
rm(reg)
rm(nor)
```

#### NDVI data
All NDVI data are calculated mean NDVI values for each respective NiN polygon from each available Sentinel-2, MODIS, or Landsat image in Google Earth Engine (GEE).

```{r, echo = F}
# downloading the unprocessed NDVI data from a data-cache et the NINA server
df.s <- readRDS("P:/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/ndvi.Sentinel.raw.RDS")
df.m <- readRDS("P:/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/ndvi.Modis.raw.RDS")
df.l <- readRDS("P:/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/ndvi.Landsat.raw.RDS")
```

##### Sentinel-2 NDVI data
The GEE code can be seen here: https://code.earthengine.google.com/2ceb0c3e03adade9e6f6d0903184b8c4
The image collection contains Sentinel imagery from June, July and August 2015-2022
To not exceed GEE memory limits, the exported files had to be iterated over a grid which resulted in 42 separate csv files. This script merges them and then merges the dataframe to the NiN data.

```{r, echo = T, eval = F}
# Import Sentinel NDVI Data
df.s <- list.files("P:/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/NDVI_data_Sentinel/", pattern = "*.csv", full.names = TRUE) %>%
  map_df(~ fread(.))
```

@@

```{r, echo = T, eval = F}
# Import MODIS NDVI Data
# MODIS NDVI is scaled by 0.0001. Mean must be divided by 10000.
df.m <- read.csv("P:\\41201785_okologisk_tilstand_2022_2023\\data\\NDVI_åpenlavland\\NDVI_data_MODIS\\modis_ndvi_ts_2000_2022.csv", )
df.m$mean <- df.m$mean / 10000
```

##### Landsat NDVI data
The GEE code can be seen here: https://code.earthengine.google.com/da8a9279238ef26d14be08a43788b6b7
The image collection contains Landsat imagery from June, July and August 1984-2022
To not exceed GEE memory limits, the exported files had to be iterated over a grid which resulted in 42 separate csv files. This script merges them and then merges the dataframe to the NiN data.

```{r, echo = T, eval = F}
# Import Landsat NDVI Data
# Set up conditional file paths
dir <- substr(getwd(), 1, 2)

path <- ifelse(dir == "C:",
  "R:/GeoSpatialData/Habitats_biotopes/Norway_Miljodirektoratet_Naturtyper_nin/Original/versjon20221231/Natur_Naturtyper_nin_norge_med_svalbard_25833/Natur_Naturtyper_NiN_norge_med_svalbard_25833.gdb",
  "/data/R/GeoSpatialData/Habitats_biotopes/Norway_Miljodirektoratet_Naturtyper_nin/Original/versjon20221231/Natur_Naturtyper_nin_norge_med_svalbard_25833/Natur_Naturtyper_NiN_norge_med_svalbard_25833.gdb"
)

pData <- ifelse(dir == "C:",
  "P:/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/NDVI_data_Landsat",
  "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/NDVI_data_Landsat"
)

# Fread doesn't like the weird path to the server version of the P drive
# hence this horrendous work around
files <- list.files(pData, pattern = "*.csv", full.names = TRUE)

df_list <- list() # initialise a list of dataframes
# read in a dataframe in each slot of the df_list
for (i in files) {
  name <- gsub("-", ".", i)
  name <- gsub(".csv", "", name)
  i <- paste(i, sep = "")
  df_list[[i]] <- assign(name, read.csv(i, header = TRUE))
}

df.l <- bind_rows(df_list, .id = "column_label")
```

### Data handling
#### NiN data
```{r, echo = T, eval = F}
# fixing variable- and ecosystem-names with special characters
colnames(nin)
colnames(nin)[c(3, 8, 17, 26, 31, 33, 34)] <- c(
  "hovedoekosystem", "kartleggingsaar", "noyaktighet",
  "omraadenavn", "uk_naertruet", "uk_sentraloekosystemfunksjon",
  "uk_spesieltdaarligkartlagt"
)
unique(nin$hovedoekosystem)

nin <- nin %>%
  mutate(hovedoekosystem = recode(hovedoekosystem,
    "Våtmark" = "Vaatmark",
    "Semi-naturlig mark" = "Semi_naturlig",
    "Naturlig åpne områder i lavlandet" = "Naturlig-aapent",
    "Naturlig åpne områder under skoggrensa" = "Naturlig_aapent"
  )) %>%
  mutate(validGeo = st_is_valid(SHAPE))

@@

## filter out only wetland data
nin.wetland <- nin %>%
  filter(hovedoekosystem %in% c("Vaatmark")) %>%
  mutate(id = identifikasjon_lokalid) %>%
  filter(validGeo) %>%
  drop_na(tilstand) %>%
  dplyr::select(id, hovedoekosystem, hovedtype, ninkartleggingsenheter, lokalitetskvalitet, tilstand, tilstand_num, ninbeskrivelsesvariabler, kartleggingsaar)

# merge NiN-data with region
nin.wetland <- st_join(nin.wetland, regnor, left = TRUE)
nin.wetland

colnames(nin.wetland)[c(1, 10)] <- c("id", "region_id")

# check that every nin-polygon still occurs only once
summary(as.factor(nin.wetland$id))
nin.wetland[nin.wetland$id == "NINFP2110008060", ]
# this one was assigned to both Western and Central Norway, we drop the latter
nin.wetland <- nin.wetland[!row.names(nin.wetland) %in% "1144.1", ]

nin.wetland <- nin.wetland %>%
  mutate(area_meters_nin = st_area(nin.wetland))
# check and edit the order of regions
levels(nin.wetland$region)
nin.wetland$region <- as.factor(nin.wetland$region)
levels(nin.wetland$region)
nin.wetland$region <- factor(nin.wetland$region, levels = c("Northern Norway", "Central Norway", "Eastern Norway", "Western Norway", "Southern Norway"))
levels(nin.wetland$region)

# splitting content in the ninbeskrivelsesvariabler-column into separate rows per variable and then separating that column into two for the variable name and value
nin.wetland2 <- nin.wetland %>%
  separate_rows(ninbeskrivelsesvariabler, sep = ",") %>%
  separate(
    col = ninbeskrivelsesvariabler,
    into = c("NiN_variable_code", "NiN_variable_value"),
    sep = "_",
    remove = F
  ) %>%
  mutate(NiN_variable_value = as.numeric(NiN_variable_value)) %>%
  mutate(NiN_variable_code = as.factor(NiN_variable_code)) %>%
  subset(select = -ninbeskrivelsesvariabler)
nin.wetland2 <- as.data.frame(nin.wetland2)
```

```{r, echo = F}
# load processed NiN-data from cache
nin.wetland <- readRDS(paste0(here::here(), "/data/cache/nin.wetland.RDS"))
nin.wetland2 <- readRDS(paste0(here::here(), "/data/cache/nin.wetland2.RDS"))
```

#### NDVI data
```{r, results='hide'}
## Sentinel-2
# join nin.wetland & Sentinel NDVI data
SentinelNDVI.wetland <- full_join(nin.wetland, df.s, by = "id")
# summary(SentinelNDVI.wetland)
SentinelNDVI.wetland <- SentinelNDVI.wetland %>%
  mutate(
    hovedoekosystem = as.factor(hovedoekosystem),
    hovedtype = as.factor(hovedtype),
    ninkartleggingsenheter = as.factor(ninkartleggingsenheter),
    lokalitetskvalitet = as.factor(lokalitetskvalitet),
    tilstand = as.factor(tilstand),
    area_meters = st_area(SentinelNDVI.wetland)
  )
# summary(SentinelNDVI.wetland)
# get rid of NAs (i.e. NDVI cells that were not in wetland polygons)
SentinelNDVI.wetland <- SentinelNDVI.wetland %>% filter(!is.na(hovedtype))
SentinelNDVI.wetland <- SentinelNDVI.wetland %>% filter(!is.na(mean))
# summary(SentinelNDVI.wetland)
# get rid of any nin-polygons smaller than the Sentinel grid cell size (100 sqm)
dim(SentinelNDVI.wetland)
SentinelNDVI.wetland <- SentinelNDVI.wetland %>% filter(as.numeric(area_meters) >= 100)
dim(SentinelNDVI.wetland)
# split date into year, month & day
SentinelNDVI.wetland <- SentinelNDVI.wetland %>%
  dplyr::mutate(
    year = lubridate::year(date),
    month = lubridate::month(date),
    day = lubridate::day(date)
  )
# add column for sub-ecosystem types
SentinelNDVI.wetland <- SentinelNDVI.wetland %>% mutate(
  subtype = substring(ninkartleggingsenheter, 4),
  subtype = str_remove(subtype, "-"),
  subtype = str_remove(subtype, "-")
)
# we are using max NDVI per year in every NiN polygon
SentinelNDVI.wetland <- SentinelNDVI.wetland %>%
  group_by(id, year) %>%
  filter(mean == max(mean, na.rm = TRUE))

@@

## MODIS
# join nin.wetland and Modis NDVI data
ModisNDVI.wetland <- full_join(nin.wetland, df.m, by = "id")

# summary(ModisNDVI.wetland)
ModisNDVI.wetland <- ModisNDVI.wetland %>%
  mutate(
    hovedoekosystem = as.factor(hovedoekosystem),
    hovedtype = as.factor(hovedtype),
    ninkartleggingsenheter = as.factor(ninkartleggingsenheter),
    lokalitetskvalitet = as.factor(lokalitetskvalitet),
    tilstand = as.factor(tilstand),
    area_meters = st_area(ModisNDVI.wetland)
  )
# summary(ModisNDVI.wetland)
# get rid of NAs (i.e. NDVI cells that were not in wetland polygons)
ModisNDVI.wetland <- ModisNDVI.wetland %>% filter(!is.na(hovedtype))
ModisNDVI.wetland <- ModisNDVI.wetland %>% filter(!is.na(mean))
# summary(ModisNDVI.wetland)
# get rid of any nin-polygons smaller than the Modis grid cell size (62500 sqm)
dim(ModisNDVI.wetland)
ModisNDVI.wetland <- ModisNDVI.wetland %>% filter(as.numeric(area_meters) >= 62500)
dim(ModisNDVI.wetland)
# split date into year, month & day
ModisNDVI.wetland <- ModisNDVI.wetland %>%
  dplyr::mutate(
    year = lubridate::year(date),
    month = lubridate::month(date),
    day = lubridate::day(date)
  )
# we are using max NDVI per year in every NiN polygon
ModisNDVI.wetland <- ModisNDVI.wetland %>%
  group_by(id, year) %>%
  filter(mean == max(mean, na.rm = TRUE))
# summary(ModisNDVI.wetland)

## Landsat
# join nin.wetland & Landsat NDVI data
LandsatNDVI.wetland <- full_join(nin.wetland, df.l, by = "id")

# summary(LandsatNDVI.wetland)
LandsatNDVI.wetland <- LandsatNDVI.wetland %>%
  mutate(
    hovedoekosystem = as.factor(hovedoekosystem),
    hovedtype = as.factor(hovedtype),
    ninkartleggingsenheter = as.factor(ninkartleggingsenheter),
    lokalitetskvalitet = as.factor(lokalitetskvalitet),
    tilstand = as.factor(tilstand),
    area_meters = st_area(LandsatNDVI.wetland)
  )
# summary(LandsatNDVI.wetland)
# get rid of NAs (i.e. NDVI cells that were not in wetland polygons)
LandsatNDVI.wetland <- LandsatNDVI.wetland %>% filter(!is.na(hovedtype))
LandsatNDVI.wetland <- LandsatNDVI.wetland %>% filter(!is.na(mean))
# summary(LandsatNDVI.wetland)
# get rid of any nin-polygons smaller than the Landsat grid cell size (900 sqm)
dim(LandsatNDVI.wetland)
LandsatNDVI.wetland <- LandsatNDVI.wetland %>% filter(as.numeric(area_meters) >= 900)
dim(LandsatNDVI.wetland)
# split date into year, month & day
LandsatNDVI.wetland <- LandsatNDVI.wetland %>%
  dplyr::mutate(
    year = lubridate::year(date),
    month = lubridate::month(date),
    day = lubridate::day(date)
  )
# we are using max NDVI per year in every NiN polygon
LandsatNDVI.wetland <- LandsatNDVI.wetland %>%
  group_by(id, year) %>%
  filter(mean == max(mean, na.rm = TRUE))

# summary(LandsatNDVI.wetland)
```

@@

# Where are the wetland sites?
```{r}
tm_shape(regnor) +
  tm_fill("GID_0", labels = "", title = "", legend.show = FALSE) +
  tm_borders() +
  tm_shape(SentinelNDVI.wetland) +
  tm_dots("hovedtype", midpoint = NA, palette = tmaptools::get_brewer_pal("YlOrRd", 6, plot = FALSE), scale = 2.5, legend.show = FALSE) +
  tm_add_legend(
    type = "symbol",
    col = tmaptools::get_brewer_pal("YlOrRd", 6, plot = FALSE),
    labels = c(
      "V1 - minerotrophic mires",
      "V3 - ombrotrophic bogs",
      "V4 - cold springs",
      "V6 - wet snowbeds",
      "V9 - semi-natural mires",
      "V10 - semi-natural wet meadows"
    ),
    title = "Major ecosystem type",
    size = 3
  ) +
  tm_layout(
    main.title = "Wetland ecosystems (NiN, Miljødirektoratets instruks)",
    main.title.size = 1.2, legend.position = c("right", "bottom"),
    legend.text.size = 1.3, legend.title.size = 1.4
  )
```

# How are polygon sizes distributed?
```{r}
summary(SentinelNDVI.wetland$area_meters)
hist(SentinelNDVI.wetland$area_meters, xlim = c(0, 20000), breaks = 50000)
abline(v = 100, lty = 2)
```
Wetland polygons stretch a large number of sizes from just above 100 sqm to more than 4 million sqm. 75% of all polygons are below 14264 sqm.

# How does NDVI vary over the years
```{r}
SentinelNDVI.wetland %>%
  ggplot(aes(x = year, y = mean)) +
  geom_point() +
  facet_grid(tilstand ~ hovedtype)
```
2022 stands out with the highest NDVI values missing.
WE omit 2022 for this analysis
```{r}
SentinelNDVI.wetland <- SentinelNDVI.wetland %>% filter(year != "2022")
```

# How do NDVI values vary between major and basic ecosystem types for sites in good condition? (using only NDVI data matching NiN-mapping years)
```{r}
SentinelNDVI.wetland %>%
  filter(tilstand == "God") %>%
  filter(year == kartleggingsaar) %>%
  ggplot(aes(x = hovedtype, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey")

SentinelNDVI.wetland %>%
  filter(tilstand == "God") %>%
  filter(year == kartleggingsaar) %>%
  ggplot(aes(x = subtype, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey") +
  facet_wrap(~hovedtype) +
  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2))
```
NDVI varies between major ecosystem types with the highest values in semi-natural wet-meadows (V10), and lowest values in ombrotrophic bogs (V3). The overlaps between the ecosystem types are considerable.
With respect to basic ecosystem types within the major types, NDVI appears to increase with increasing limestone level (C1-C4, C5-C8, and E1-E3 in V1, C1-C3 in the other major types), and also here the overlap is large.

# How does NDVI vary across regions in wetlands (major types) in good condition? (using only NDVI data matching NiN-mapping years)
```{r}
SentinelNDVI.wetland %>%
  filter(tilstand == "God") %>%
  filter(year == kartleggingsaar) %>%
  filter(!is.na(region)) %>%
  ggplot(aes(x = region, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey") +
  facet_wrap(~hovedtype) +
  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2))
```
There is no obvious trends in geography.
